---
title: "Chapter 2: Basic Principles of Survival Analysis"
subtitle: "Understanding Hazard and Survival Functions"
author: "Eric Delmelle"
date: today
format: 
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    code-fold: false
    code-tools: true
    df-print: paged
    number-sections: true
    number-offset: 1
  pdf:
    toc: true
    number-sections: true
    number-offset: 1
    geometry: margin=1in
execute:
  warning: false
  message: false
  fig-width: 8
  fig-height: 5
---

## Learning Objectives

- Define and interpret survival and hazard functions
- Calculate survival probabilities from hazard functions
- Work with parametric survival distributions (exponential, Weibull, gamma)
- Compute mean and median survival times
- Apply maximum likelihood estimation to simple survival problems

## Setup

```{r setup}
#| label: setup

library(survival)
library(ggplot2)
library(dplyr)
library(gridExtra)

# Set random seed for reproducibility
set.seed(123)
```

---

## The Hazard and Survival Functions

### Key Definitions

The **survival function** defines the probability of surviving up to time $t$:

$$S(t) = P(T > t), \quad 0 < t < \infty$$

- Takes value 1 at time 0
- Decreases (or remains constant) over time  
- Never drops below 0
- Right continuous

The **hazard function** is the instantaneous failure rate:

$$h(t) = \lim_{\delta \to 0} \frac{P(t < T < t + \delta | T > t)}{\delta}$$

- Also called the intensity function or force of mortality
- Probability of failure in the next small interval, given survival to time $t$

### Key Relationships

The hazard and survival functions are related by:

$$h(t) = \frac{f(t)}{S(t)}$$

where $f(t)$ is the probability density function.

The **cumulative hazard function** is:

$$H(t) = \int_0^t h(u) du$$

And the survival function can be expressed as:

$$S(t) = \exp(-H(t))$$

---

## Light Bulb Demonstration: Survival Analysis in Action

### Exercise 0: Physical Demonstration (20 minutes)

**Materials:** 5 light bulbs with failure times pre-marked

Let's make survival analysis concrete with a physical demonstration.

#### Part A: Data Collection (5 minutes)
Reveal the failure times written on each bulb:
```
Bulb A: Failed at ___ hours 
Bulb B: Failed at ___ hours   
Bulb C: Censored at ___ hours (still working when removed)
Bulb D: Failed at ___ hours 
Bulb E: Censored at ___ hours (still working when removed)
```

#### Part B: Manual S(t) Construction (10 minutes)

**Step 1:** Organize events chronologically
```
Time ___: Bulb ___ fails
Time ___: Bulb ___ fails  
Time ___: Bulb ___ censored
Time ___: Bulb ___ fails
Time ___: Bulb ___ censored
```

**Step 2:** Build survival table by hand

| Time | At Risk | Deaths | Censored | Conditional Survival | S(t) |
|------|---------|--------|----------|---------------------|------|
| 0    | 5       | 0      | 0        | 1.0                | 1.0  |
| ___  | ___     | ___    | ___      | ___/___            | ___ |
| ___  | ___     | ___    | ___      | ___/___            | ___ |
| ___  | ___     | ___    | ___      | ___/___            | ___ |
| ___  | ___     | ___    | ___      | ___/___            | ___ |
| ___  | ___     | ___    | ___      | ___/___            | ___ |

**Step 3:** Draw survival curve on blackboard

-   Plot time on x-axis, S(t) on y-axis
-   Show step function dropping only at death times
-   Mark censoring points with tick marks

#### Part C: R Visualization (5 minutes)

```{r lightbulb-demo}
#| label: lightbulb-demo
#| fig-height: 6

# Input actual class data here
bulb_times <- c(3, 7, 10, 12, 15)  # Replace with actual times
bulb_status <- c(1, 1, 0, 1, 0)    # 1=failed, 0=censored

# Manual calculation verification
cat("Manual Calculations:\n")
cat("Bulb-hours:", sum(bulb_times), "\n")
cat("Failures:", sum(bulb_status), "\n")
cat("Failure rate:", round(sum(bulb_status)/sum(bulb_times), 4), "per hour\n\n")

# Kaplan-Meier curve
bulb_surv <- Surv(bulb_times, bulb_status)
km_fit <- survfit(bulb_surv ~ 1)

# Create detailed plot
par(mfrow=c(1,2))

# Plot 1: Basic survival curve
plot(km_fit, xlab="Time (hours)", ylab="Survival Probability", 
     main="Light Bulb Survival", conf.int=FALSE, lwd=2)
grid()

# Add censoring marks
censor_times <- bulb_times[bulb_status == 0]
if(length(censor_times) > 0) {
  censor_surv <- summary(km_fit, times=censor_times)$surv
  points(censor_times, censor_surv, pch=3, cex=1.5, lwd=2)
}

# Plot 2: Step-by-step data
plot(bulb_times, 1:length(bulb_times), 
     pch=ifelse(bulb_status==1, 4, 1),
     col=ifelse(bulb_status==1, "red", "blue"),
     xlab="Bulb Time (hours)", ylab="Bulb ID",
     main="Individual Bulb Data",
     cex=1.5, lwd=2)
legend("bottomright", c("Failed", "Censored"), 
       pch=c(4,1), col=c("red", "blue"))

par(mfrow=c(1,1))

# Print survival table
cat("Kaplan-Meier Table:\n")
print(summary(km_fit))
```

---

## Building S(t) Step-by-Step: Patient Data

### Exercise 1: Manual Construction (15 minutes)

**Dataset:** 6 patients - let's build the survival curve by hand

| Patient | Time | Status | 
|---------|------|--------|
| 1       | 1    | Death  |
| 2       | 2    | Death  |
| 3       | 3    | Censored |
| 4       | 4    | Death  |
| 5       | 4    | Death  |
| 6       | 5    | Censored |

#### Step-by-Step Manual Calculation (10 minutes)

**Create survival table together:**

| Time | At Risk | Deaths | Conditional Survival | S(t) | Calculation |
|------|---------|--------|---------------------|------|-------------|
| 0    | 6       | 0      | 1.0                | 1.0  | Starting point |
| 1    | ___     | ___    | ___                | ___  | 1.0 × (___ / ___) |
| 2    | ___     | ___    | ___                | ___  | ___ × (___ / ___) |
| 3    | ___     | ___    | ___                | ___  | No deaths → same |
| 4    | ___     | ___    | ___                | ___  | ___ × (___ / ___) |
| 5    | ___     | ___    | ___                | ___  | ___ × (___ / ___) |

**Key Questions to Work Through:**
1. At time 3, patient 3 is censored. What happens to S(t)?
2. At time 4, two patients die. How do we handle this?
3. When do we stop calculating?

#### Visualization in R (5 minutes)

```{r patient-survival}
#| label: patient-survival
#| fig-height: 6

# Patient data
time <- c(1, 2, 3, 4, 4, 5)
status <- c(1, 1, 0, 1, 1, 0)

# Manual calculations first
cat("Manual Verification:\n")
cat("Person-time:", sum(time), "years\n")
cat("Deaths:", sum(status), "\n")
cat("Overall rate:", round(sum(status)/sum(time), 4), "per year\n\n")

# Survival analysis
surv_obj <- Surv(time, status)
km_patient <- survfit(surv_obj ~ 1)

# Create comprehensive visualization
par(mfrow=c(2,2))

# Plot 1: Survival curve
plot(km_patient, xlab="Time (years)", ylab="Survival Probability", 
     main="Patient Survival Curve", conf.int=FALSE, lwd=2)
grid()

# Plot 2: Risk set over time
times <- sort(unique(time[status==1]))
risk_set <- sapply(times, function(t) sum(time >= t))
plot(times, risk_set, type="s", lwd=2,
     xlab="Time", ylab="Number at Risk", 
     main="Risk Set Over Time")
grid()

# Plot 3: Individual patient data (dumbbell timeline)
plot(time, 1:6, pch=ifelse(status==1, 4, 1),
     col=ifelse(status==1, "red", "blue"),
     xlab="Time (years)", ylab="Patient Number", 
     main="Individual Patient Lifetimes",
     cex=1.5, lwd=2, xlim=c(0, max(time)*1.1))

# Add horizontal lines from 0 to each event time
for(i in 1:6) {
  segments(0, i, time[i], i, 
           col=ifelse(status[i]==1, "red", "blue"), 
           lwd=2)
}

legend("bottomright", c("Death", "Censored"), 
       pch=c(4,1), col=c("red", "blue"))

# Plot 4: Cumulative events
cum_events <- cumsum(status[order(time)])
ordered_times <- sort(time)
plot(ordered_times, cum_events, type="s", lwd=2,
     xlab="Time", ylab="Cumulative Deaths",
     main="Cumulative Events")
grid()

par(mfrow=c(1,1))

# Detailed survival table
cat("Detailed Survival Table:\n")
print(summary(km_patient))
```

---

## Exponential Model: Theory Meets Practice

### Exercise 2: By-Hand Exponential Calculations (10 minutes)

Using our patient data, let's fit an exponential model manually.

#### MLE Calculation by Hand
```
λ̂ = Deaths ÷ Person-time = ___ ÷ ___ = ___ per year
```

#### Exponential Predictions
```
Mean survival = 1/λ̂ = ___
Median survival = ln(2)/λ̂ = ___
S(3) = exp(-λ̂ × 3) = ___
```

#### Compare Models Visually

```{r exponential-comparison}
#| label: exponential-comparison
#| fig-height: 8

# Calculate MLE
d <- sum(status)
V <- sum(time) 
lambda_hat <- d/V

cat("Exponential Model:\n")
cat("λ̂ =", round(lambda_hat, 4), "per year\n")
cat("Mean survival =", round(1/lambda_hat, 2), "years\n")
cat("Median survival =", round(log(2)/lambda_hat, 2), "years\n\n")

# Create comparison plots
par(mfrow=c(2,2))

# Plot 1: Both survival curves
plot(km_patient, conf.int=FALSE, lwd=2,
     xlab="Time (years)", ylab="Survival Probability",
     main="Kaplan-Meier vs Exponential")
curve(exp(-lambda_hat * x), from=0, to=6, add=TRUE, 
      col="red", lwd=2, lty=2)
legend("topright", c("Kaplan-Meier", "Exponential"), 
       col=c("black", "red"), lwd=2, lty=c(1,2))
grid()

# Plot 2: Hazard functions
plot(c(0, max(time)), c(lambda_hat, lambda_hat), type="l",
     lwd=2, col="red", xlab="Time", ylab="Hazard Rate",
     main="Exponential Hazard (Constant)")
grid()

# Plot 3: Residuals
km_times <- summary(km_patient)$time
km_surv <- summary(km_patient)$surv
exp_surv <- exp(-lambda_hat * km_times)
plot(km_times, km_surv - exp_surv, type="p", pch=16,
     xlab="Time", ylab="KM - Exponential",
     main="Model Residuals")
abline(h=0, lty=2)
grid()

# Plot 4: Q-Q plot for exponential
theoretical_quantiles <- qexp(ppoints(length(time[status==1])), rate=lambda_hat)
observed_quantiles <- sort(time[status==1])
plot(theoretical_quantiles, observed_quantiles,
     xlab="Theoretical Quantiles", ylab="Observed Quantiles",
     main="Q-Q Plot (Exponential)")
abline(0, 1, lty=2)
grid()

par(mfrow=c(1,1))

# Summary comparison
cat("Model Comparison at t=3:\n")
km_3 <- summary(km_patient, times=3)$surv
exp_3 <- exp(-lambda_hat * 3)
cat("Kaplan-Meier S(3) =", round(km_3, 3), "\n")
cat("Exponential S(3) =", round(exp_3, 3), "\n")
cat("Difference =", round(km_3 - exp_3, 3), "\n")
```

Let's work through a small dataset by hand. Consider these 6 patients:

| Patient | Time | Status | 
|---------|------|--------|
| 1       | 7    | Death  |
| 2       | 6    | Death  |
| 3       | 6    | Censored |
| 4       | 5    | Death  |
| 5       | 2    | Censored |
| 6       | 4    | Censored |

### Exercise 1A: Basic Calculations (Do by Hand)

**Step 1:** Calculate the total person-time at risk
```
Person-time = 7 + 6 + 6 + 5 + 2 + 4 = ___ years
```

**Step 2:** Count the total number of deaths
```
Deaths = ___
```

**Step 3:** Calculate the overall hazard rate (deaths/person-time)
```
Hazard rate = Deaths / Person-time = ___ / ___ = ___
```

### Exercise 1B: Verify with R

```{r exercise-1-verify}
#| label: exercise-1-verify

# Patient data
time <- c(7, 6, 6, 5, 2, 4)
status <- c(1, 1, 0, 1, 0, 0)  # 1 = death, 0 = censored

# Calculate person-time and deaths
person_time <- sum(time)
deaths <- sum(status)

print(paste("Person-time:", person_time))
print(paste("Deaths:", deaths))
print(paste("Hazard rate:", deaths/person_time))
```

---

## Parametric Survival Distributions

### The Exponential Distribution

The simplest survival distribution has a **constant hazard**: $h(t) = \lambda$

**Key Properties:**

-   Hazard: $h(t) = \lambda$ (constant)
-   Cumulative hazard: $H(t) = \lambda t$
-   Survival: $S(t) = e^{-\lambda t}$
-   Mean survival: $E(T) = 1/\lambda$
-   Median survival: $t_{med} = \ln(2)/\lambda$

### Exercise 2: Exponential Distribution Calculations (By Hand)

Suppose we have an exponential distribution with $\lambda = 0.1$ (per year).

**Step 1:** Calculate the probability of surviving beyond 5 years
```
S(5) = e^(-λt) = e^(-0.1 × 5) = e^(-0.5) = ___
```

**Step 2:** Calculate the mean survival time
```
E(T) = 1/λ = 1/0.1 = ___ years
```

**Step 3:** Calculate the median survival time
```
t_med = ln(2)/λ = 0.693/0.1 = ___ years
```

**Step 4:** What's the hazard of death at any time t?
```
h(t) = ___ (constant)
```

### Verify Your Calculations

```{r exponential-verification}
#| label: exponential-verification

lambda <- 0.1

# Survival at 5 years
surv_5 <- exp(-lambda * 5)
print(paste("S(5) =", round(surv_5, 4)))

# Mean survival
mean_surv <- 1/lambda
print(paste("Mean survival =", mean_surv, "years"))

# Median survival
median_surv <- log(2)/lambda
print(paste("Median survival =", round(median_surv, 2), "years"))

# Verify: S(median) should equal 0.5
print(paste("S(median) =", round(exp(-lambda * median_surv), 3)))
```

### Visualizing the Exponential Distribution

```{r exponential-plots}
#| label: exponential-plots

# Create time sequence
time_seq <- seq(0, 30, 0.1)

# Calculate survival and hazard functions
surv_func <- exp(-lambda * time_seq)
hazard_func <- rep(lambda, length(time_seq))

# Create plots
p1 <- ggplot(data.frame(time = time_seq, survival = surv_func), 
             aes(x = time, y = survival)) +
  geom_line(size = 1, color = "blue") +
  labs(title = "Exponential Survival Function (λ = 0.1)",
       x = "Time (years)", y = "Survival Probability") +
  theme_minimal()

p2 <- ggplot(data.frame(time = time_seq, hazard = hazard_func), 
             aes(x = time, y = hazard)) +
  geom_line(size = 1, color = "red") +
  labs(title = "Exponential Hazard Function (λ = 0.1)",
       x = "Time (years)", y = "Hazard Rate") +
  theme_minimal()

grid.arrange(p1, p2, ncol = 2)
```

---

## The Weibull Distribution

More flexible than exponential, with hazard function:

$$h(t) = \alpha \lambda^\alpha t^{\alpha-1}$$

**Parameters:**

-   $\alpha$ = shape parameter
-   $\lambda$ = scale parameter

**Key Properties:**

-   When $\alpha = 1$: reduces to exponential
-   When $\alpha > 1$: increasing hazard
-   When $\alpha < 1$: decreasing hazard

### Exercise 3: Weibull Distribution (By Hand)

Consider a Weibull distribution with $\alpha = 2$ and $\lambda = 0.1$.

**Step 1:** What type of hazard pattern does this represent?
```
Since α = 2 > 1, the hazard is _______ over time
```

**Step 2:** Calculate the hazard at time t = 5 years
```
h(5) = α λ^α t^(α-1) = 2 × (0.1)^2 × 5^(2-1) = 2 × 0.01 × 5 = ___
```

**Step 3:** Calculate the cumulative hazard at t = 5
```
H(5) = (λt)^α = (0.1 × 5)^2 = (0.5)^2 = ___
```

**Step 4:** Calculate the survival probability at t = 5
```
S(5) = exp(-H(5)) = exp(-0.25) = ___
```

### Verify Weibull Calculations

```{r weibull-verification}
#| label: weibull-verification

alpha <- 2
lambda <- 0.1
t <- 5

# Hazard at t=5
hazard_5 <- alpha * (lambda^alpha) * (t^(alpha-1))
print(paste("h(5) =", hazard_5))

# Cumulative hazard at t=5
cum_hazard_5 <- (lambda * t)^alpha
print(paste("H(5) =", cum_hazard_5))

# Survival at t=5
surv_5_weibull <- exp(-cum_hazard_5)
print(paste("S(5) =", round(surv_5_weibull, 4)))

# Verify using R's built-in function
surv_5_r <- pweibull(5, shape=alpha, scale=1/lambda, lower.tail=FALSE)
print(paste("S(5) using R =", round(surv_5_r, 4)))
```

### Comparing Different Weibull Shapes

```{r weibull-comparison}
#| label: weibull-comparison

# Define parameters for comparison
time_seq <- seq(0, 20, 0.1)
lambda <- 0.1

# Different shape parameters
alphas <- c(0.75, 1, 1.5, 2)

# Create data frame for plotting
weibull_data <- data.frame()

for(alpha in alphas) {
  hazard <- alpha * (lambda^alpha) * (time_seq^(alpha-1))
  survival <- exp(-(lambda * time_seq)^alpha)
  
  temp_data <- data.frame(
    time = time_seq,
    hazard = hazard,
    survival = survival,
    alpha = as.factor(alpha)
  )
  weibull_data <- rbind(weibull_data, temp_data)
}

# Plot hazard functions
p1 <- ggplot(weibull_data, aes(x = time, y = hazard, color = alpha)) +
  geom_line(size = 1) +
  labs(title = "Weibull Hazard Functions",
       x = "Time", y = "Hazard Rate",
       color = "Shape (α)") +
  theme_minimal() +
  ylim(0, 0.3)

# Plot survival functions  
p2 <- ggplot(weibull_data, aes(x = time, y = survival, color = alpha)) +
  geom_line(size = 1) +
  labs(title = "Weibull Survival Functions",
       x = "Time", y = "Survival Probability",
       color = "Shape (α)") +
  theme_minimal()

grid.arrange(p1, p2, ncol = 2)
```

---

## Maximum Likelihood Estimation

### The Likelihood Function

For survival data with censoring, the likelihood is:

$$L(\theta) = \prod_{i=1}^n h(t_i; \theta)^{\delta_i} S(t_i; \theta)$$

where $\delta_i = 1$ for deaths and $\delta_i = 0$ for censored observations.

### Exercise 4: MLE for Exponential Distribution (By Hand)

Using our 6-patient dataset from Exercise 1:

**The log-likelihood for exponential distribution is:**
$$\ell(\lambda) = d \log(\lambda) - \lambda V$$

where $d$ = number of deaths and $V$ = total person-time.

**Step 1:** Write the log-likelihood for our data
```
d = ___ deaths
V = ___ person-years
ℓ(λ) = ___ log(λ) - λ × ___
```

**Step 2:** Find the MLE by taking the derivative and setting to zero
```
dℓ/dλ = d/λ - V = 0
Solving: λ̂ = d/V = ___ / ___ = ___
```

**Step 3:** Calculate the variance of the estimate
```
Var(λ̂) ≈ λ̂²/d = (_____)²/___ = ___
Standard Error = √Var(λ̂) = ___
```

### Verify MLE Calculation

```{r mle-verification}
#| label: mle-verification

# Our data
time <- c(7, 6, 6, 5, 2, 4)
status <- c(1, 1, 0, 1, 0, 0)

d <- sum(status)  # deaths
V <- sum(time)    # person-time

# MLE
lambda_hat <- d/V
print(paste("λ̂ =", round(lambda_hat, 4)))

# Variance and standard error
var_lambda <- lambda_hat^2 / d
se_lambda <- sqrt(var_lambda)

print(paste("Variance of λ̂ =", round(var_lambda, 6)))
print(paste("Standard Error =", round(se_lambda, 4)))

# 95% Confidence interval (using normal approximation)
ci_lower <- lambda_hat - 1.96 * se_lambda
ci_upper <- lambda_hat + 1.96 * se_lambda

print(paste("95% CI: (", round(ci_lower, 4), ",", round(ci_upper, 4), ")"))
```

---

## Hands-On Workshop Exercise

### Exercise 5: Complete Analysis by Hand, Then Verify

You are given this survival dataset:

| Patient | Time | Status |
|---------|------|--------|
| A       | 12   | Death  |
| B       | 8    | Death  |
| C       | 15   | Censored |
| D       | 10   | Death  |
| E       | 6    | Censored |
| F       | 9    | Death  |

**Part A: Basic Statistics (Work by hand first)**

1. Total person-time: ___________
2. Number of deaths: ___________  
3. MLE for exponential λ: ___________
4. Mean survival time: ___________
5. Median survival time: ___________

**Part B: Probability Calculations**

Assuming exponential distribution with your estimated λ:

1. P(survive > 10 years) = ___________
2. P(die between years 5-10) = ___________
3. Hazard rate at any time = ___________

**Part C: Verify Your Work**

```{r workshop-exercise}
#| label: workshop-exercise
#| eval: false

# Fill in the values
time_workshop <- c(__, __, __, __, __, __)
status_workshop <- c(__, __, __, __, __, __)  # 1=death, 0=censored

# Calculate basic statistics
person_time_total <- sum(time_workshop)
deaths_total <- sum(status_workshop)
lambda_est <- deaths_total / person_time_total

print(paste("Person-time:", person_time_total))
print(paste("Deaths:", deaths_total))
print(paste("λ estimate:", round(lambda_est, 4)))
print(paste("Mean survival:", round(1/lambda_est, 2)))
print(paste("Median survival:", round(log(2)/lambda_est, 2)))

# Probability calculations
prob_survive_10 <- exp(-lambda_est * 10)
prob_die_5_to_10 <- exp(-lambda_est * 5) - exp(-lambda_est * 10)

print(paste("P(survive > 10):", round(prob_survive_10, 4)))
print(paste("P(die between 5-10):", round(prob_die_5_to_10, 4)))
```

### Solution

```{r workshop-solution}
#| label: workshop-solution

# Solution
time_workshop <- c(12, 8, 15, 10, 6, 9)
status_workshop <- c(1, 1, 0, 1, 0, 1)

person_time_total <- sum(time_workshop)
deaths_total <- sum(status_workshop)
lambda_est <- deaths_total / person_time_total

print("=== SOLUTION ===")
print(paste("Person-time:", person_time_total, "years"))
print(paste("Deaths:", deaths_total))
print(paste("λ estimate:", round(lambda_est, 4)))
print(paste("Mean survival:", round(1/lambda_est, 2), "years"))
print(paste("Median survival:", round(log(2)/lambda_est, 2), "years"))

# Probability calculations
prob_survive_10 <- exp(-lambda_est * 10)
prob_die_5_to_10 <- exp(-lambda_est * 5) - exp(-lambda_est * 10)

print(paste("P(survive > 10 years):", round(prob_survive_10, 4)))
print(paste("P(die between 5-10 years):", round(prob_die_5_to_10, 4)))
print(paste("Hazard rate:", round(lambda_est, 4), "(constant)"))
```

---

## Summary and Key Takeaways

### Essential Formulas to Remember

**Exponential Distribution:**
-   Survival: $S(t) = e^{-\lambda t}$  
-   Hazard: $h(t) = \lambda$ (constant)
-   Mean: $1/\lambda$
-   Median: $\ln(2)/\lambda$
-   MLE: $\hat{\lambda} = d/V$

**General Relationships:**
-   $S(t) = \exp(-H(t))$ where $H(t) = \int_0^t h(u)du$
-   $h(t) = f(t)/S(t)$
-   Mean survival = $\int_0^\infty S(t)dt$


---

## Additional Practice Problems

### Problem 1
If the hazard rate is constant at 0.05 per year, what percentage of patients survive beyond 20 years?

### Problem 2  
A study reports a median survival of 8 years. Assuming exponential distribution, what is the hazard rate?

### Problem 3
Given survival times: 3, 5, 7+, 9, 12+ (+ indicates censoring), calculate the MLE for λ.

---

## References

Moore, D. *Applied Survival Analysis Using R*. Chapter 2: Basic Principles of Survival Analysis.

---

## Session Information

```{r session-info}
#| label: session-info

sessionInfo()
```

[← Return to Course Materials](https://bsta-150.github.io/course/)

[Link to qmd (quarto markdown)](https://github.com/BSTA-150/chapter2/blob/main/chapter2.qmd)